---
title: "01. Loading and merging data sets"
execute: 
  echo: false
  eval: true
format: html
editor: 
  markdown: 
    wrap: 80
---

## Loading libraries

```{r}
library(tidyverse)
```

## Source of data

The data set used in this project was originally published by [Gloub *et al.*
(1999)](https://www-science-org.proxy.findit.cvt.dk/doi/10.1126/science.286.5439.531#body-ref-RF14)
and was downloaded from
<https://www.kaggle.com/datasets/crawford/gene-expression/>.

## Brief data description

There are 3 files of interest in `../_raw/`:

1.  `actual.csv`,

2.  `data_set_ALL_AML_independent.csv`, and

3.  `data_set_ALL_AML_train.csv`.

The first one contains (numeric) patient IDs in column 'patient' and cancer type
(ALL for acute lymphoblastic leukemia, AML for acute myeloid leukemia) in column
'cancer'. The remaining two contain gene expression data per gene in row, while
the columns hold information on each gene's description, its accession number,
patient ID and something denoted gene "call". The latter refers to whether the
authors of the paper deemed the gene present (P), marginal (M), or absent (A) in
the sample from each patient.

## Loading data

```{r}
# Reading each of the 3 data files
patient_data <- read_csv("../_raw/actual.csv")
expression_indep <- read_csv("../_raw/data_set_ALL_AML_independent.csv")
expression_train <- read_csv("../_raw/data_set_ALL_AML_train.csv")

# Combining the 2 tibbles with gene expression data
expression <- merge(x = expression_indep,
                    y = expression_train,
                    by.x = c("Gene Description",
                             "Gene Accession Number"),
                    by.y = c("Gene Description",
                             "Gene Accession Number")
                    )
```

## Writing combined data

```{r}
write.csv(expression,
          "../data/01_expression_data.csv",
          row.names = FALSE
          )

write.csv(patient_data,
          "../data/01_patient_data.csv",
          row.names = FALSE
          )
 
# Removing objects to free memory
remove(expression_indep,
       expression_train,
       expression,
       patient_data
       )
```